Integration of Multiple Constraints in ACG

### 2013-09-21 lenls/paper.tex
LENLS: Wrote up the example grammar

\begin{align*}
N_{de_1} &: NP\_NEG{=}F \limp N\_NEG{=}F \limp N\_NEG{=}F \\
N_{de_2} &: NP\_NEG{=}F \limp N\_NEG{=}T \limp N\_NEG{=}T \\
N_{de_3} &: NP\_NEG{=}T \limp N\_NEG{=}F \limp N\_NEG{=}T
\end{align*}

In the above, we elaborate the types $NP$ and $N$ with features ($\_NEG{=}F$
and $\_NEG{=}T$) and we give the different types for the preposition \emph{de}
in the fragment for negation ($N$). The types accept an $NP$ and an $N$ as
arguments with any combination of values for the feature $NEG$, except for the
case when both the prepositional $NP$ and the $N$ being modified both contain
free negative determiners (i.e. there is no $N_{de_4} : NP\_NEG{=}T \limp
N\_NEG{=}T \limp N\_NEG{=}\ldots$). This encodes a constraint that there can
be only one free negative determiner per phrase (free as in not hidden inside
an embedded clause). Besides this constraint, the types also dictate how the
information about negative determiners should propagate from the argument
constituent to the complex constituent (in this case, it is simple
disjunction).

@@

\begin{align*}
C_{de_{11}} :\ &(NP\_\textcolor{red}{NEG{=}T}\_\textcolor{green}{VAR{=}F}\_\textcolor{blue}{NUM{=}PL}) \\
&\limp (N\_\textcolor{red}{NEG{=}F}\_\textcolor{blue}{NUM{=}SG}) \\
&\limp (N\_\textcolor{red}{NEG{=}T}\_\textcolor{blue}{NUM{=}SG})
\end{align*}

@@

\section{Example Grammar}

Finally, we construct a G-ACG which integrates the French negation constraint
discussed in Section~\ref{sec:constraints}, the constraints on extraction
introduced in \cite{pogodalla2012controlling} and a constraint handling
agreement in a single grammar specification. The constraints are implemented
exactly as they were originally described. The underlying syntactic signature
is defined without any concern for the additional constraints and the
syntax-semantics interface is kept simple as well.

>>>>

\begin{align*}
N_{de_1} &: NP_{NEG{=}F} \limp N_{NEG{=}F} \limp N_{NEG{=}F} \\
N_{de_2} &: NP_{NEG{=}F} \limp N_{NEG{=}T} \limp N_{NEG{=}T} \\
N_{de_3} &: NP_{NEG{=}T} \limp N_{NEG{=}F} \limp N_{NEG{=}T}
\end{align*}

In the above, we elaborate the types $NP$ and $N$ with features ($_{NEG{=}F}$
and $_{NEG{=}T}$) and we give the different types for the preposition \emph{de}
in the fragment for negation ($N$). The types accept an $NP$ and an $N$ as
arguments with any combination of values for the feature $NEG$, except for the
case when both the prepositional $NP$ and the $N$ being modified both contain
free negative determiners (i.e. there is no $N_{de_4} : NP_{NEG{=}T} \limp
N_{NEG{=}T} \limp N\_NEG{=}\ldots$). This encodes a constraint that there can
be only one free negative determiner per phrase (free as in not hidden inside
an embedded clause). Besides this constraint, the types also dictate how the
information about negative determiners should propagate from the argument
constituent to the complex constituent (in this case, it is simple
disjunction).

@@

$$
C_{de_{11}} :\ NP_{\textcolor{red}{NEG{=}T}, \textcolor{green}{VAR{=}F}, \textcolor{blue}{NUM{=}PL}}
\limp N_{\textcolor{red}{NEG{=}F},\textcolor{blue}{NUM{=}SG}}
\limp N_{\textcolor{red}{NEG{=}T},\textcolor{blue}{NUM{=}SG}}
$$

@@

\section{Illustration}

In this final section, we assemble a G-ACG which integrates the French
negation constraint discussed in Section~\ref{sec:constraints}, the
constraints on extraction introduced in \cite{pogodalla2012controlling} and a
constraint handling agreement in a single grammar specification to demonstrate
our approach.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/single-cont.pdf}
    \caption{\label{fig:single-cont} An extraction constraint }
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/merged-conts.pdf}
    \caption{\label{fig:merged-conts} G-ACG merging the three extraction
      constraints.}
  \end{subfigure}
  \caption{\label{fig:conts} Combining the constraints on extraction.}
\end{figure}

We start with the extraction constraints. \cite{pogodalla2012controlling}
describes three different constraints on extraction, all of them expressed
using the Constraint pattern (Figure~\ref{fig:single-cont}). We can take the
union of these three G-ACGs to get the G-ACG on Figure~\ref{fig:merged-conts}.
Next, we can incorporate knowledge from another paper,
\cite{pogodalla2007generalizing}, by splitting the $\mathcal{L}_{Syn}$ lexicon
and introducing a new intermediate signature
(Figure~\ref{fig:conts-elab}). The new signature, $Syntax$, deals purely with
syntax without any issues of operator scope and its functions have lower-order
types. We will finally add our contribution, the signatures $Neg$ and $Agr$
and the lexicons $\mathcal{L}_{Neg}$ and $\mathcal{L}_{Agr}$ which implement
the negation phenomenon we discussed before and some simple notion of
agreement, respectively.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/conts-elab.pdf}
    \caption{\label{fig:conts-elab} Elaborating the lexicon from syntax to
      strings. }
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/final-synth.pdf}
    \caption{\label{fig:final-synth} Adding our own syntactic constraints,
      $Neg$ and $Agr$.}
  \end{subfigure}
  \caption{\label{fig:final} Building the example grammar up to its final
    state.}
\end{figure}

We notice that while the three constraints introduced in
\cite{pogodalla2012controlling} are expressed in terms of the $Syn$ signature,
our constraints $Agr$ and $Neg$ use a different interface, the $Syntax$
signature, to constrain the languages being defined. The $Cont_i$ constraints
are not actually constraining syntax itself, but the syntax-semantics
interface. This is in some cases a deliberate decision, since e.g. $Cont_1$ is
the constraint enforcing that quantifier scope must not escape outside of
tensed clauses. The $Syntax$ signature is not equipped to talk about operator
scope and so the constraint is expressed above the $Syn$ signature. On the
other hand, constraints $Cont_2$ and $Cont_3$ from
\cite{pogodalla2012controlling} are purely syntactic. However, we see it as a
perk of our approach that we can ignore that and import the constraint
wholesale without having to adapt them.

\subsection{The $Neg$ Signature}

We will now look under the covers of the two signatures implementing the
constraints, starting with negation and the $Neg$ signature.

As was said in Section~\ref{sec:constraints}, we will proceed by splitting the
atomic types $N$ and $NP$ into $N_{NEG{=}F}$ and $N_{NEG{=}T}$, and
$NP_{NEG{=}F}$ and $NP_{NEG{=}T}$, respectively. We will use these new types to
distinguish whether an $N$ or $NP$ phrase contains a negative determiner that
can pair up with a negated verb. Where before we had a single type, we will
now have possibly multiple to account for the different values of the
arguments' $NEG$ features. Here are some representative examples:

\begin{align*}
N_{aucun} &: N_{NEG{=}F} \limp NP_{NEG{=}T} \\
N_{le_1} &: N_{NEG{=}F} \limp NP_{NEG{=}F} \\
N_{le_2} &: N_{NEG{=}T} \limp NP_{NEG{=}T} \\
N_{ne^{tv}_1} &: (NP_{NEG{=}F} \limp NP_{NEG{=}F} \limp S) \limp (NP_{NEG{=}T} \limp NP_{NEG{=}F} \limp S) \\
N_{ne^{tv}_2} &: (NP_{NEG{=}F} \limp NP_{NEG{=}F} \limp S) \limp (NP_{NEG{=}F}
\limp NP_{NEG{=}T} \limp S) \\
N_{ne^{tv}_3} &: (NP_{NEG{=}F} \limp NP_{NEG{=}F} \limp S) \limp (NP_{NEG{=}T}
\limp NP_{NEG{=}T} \limp S) \\
N_{aime} &: NP_{NEG{=}F} \limp NP_{NEG{=}F} \limp S \\
N_{que_1} &: (NP_{NEG{=}F} \limp S) \limp N_{NEG{=}F} \limp N_{NEG{=}F} \\
N_{que_2} &: (NP_{NEG{=}F} \limp S) \limp N_{NEG{=}T} \limp N_{NEG{=}T} \\
\end{align*}

We see that the negative determiner \emph{aucun} can only attach to phrases
which do not already contain a free negative determiner and that the resulting
phrase is marked as having a free negative determiner. On the other hand, the
determiner \emph{le} does not have any interaction and is still able to take
any $N$ phrase, preserving any free negative determiners inside. The type of
the negative particle \emph{ne} modifies a verb (in the example above,
$N_{ne^{tv}_i}$, a transitive verb) and produces a new verb that makes sure
that at least one of its arguments contains a negative determiner. Verbs are
not agnostic about the $NEG$ feature. By default, they specifically demand it
to be false, since a negative determiner cannot be coupled with a non-negated
verb. Finally, the types of the relativizer \emph{que} tell us two other
things: the $NP$ trace is considered as containing no negative determiners,
and a relative clause does not care about or alter the presence of free
negative determiners in the $N$ it modifies.

The lexicon which translates from this signature to $Syntax$ is a
straightforward relabeling. Its items can be characterized by the following
schema:

$$
\mathcal{L}_{Neg}(N_{wordform_{i}}) = C_{wordform}
$$

where $C_{wordform}$ is the constant in $Syntax$ corresponding to the
$wordform$.

\subsection{The $Agr$ Signature}

The $Agr$ signature will be constructed using the same strategy as $Neg$. $N$
and $NP$ will be subdivided into $N_{NUM{=}SG}$, $N_{NUM{=}PL}$,
$NP_{NUM{=}SG}$ and $NP_{NUM{=}PL}$ (we only treat number agreement in this
example). Here are some example types:

\begin{align*}
A_{tatou} &: N_{NUM{=}SG} \\
A_{tatous} &: N_{NUM{=}PL} \\
A_{le} &: N_{NUM{=}SG} \limp NP_{NUM{=}SG} \\
A_{les} &: N_{NUM{=}PL} \limp NP_{NUM{=}PL} \\
A_{aime_1} &: NP_{NUM{=}SG} \limp NP_{NUM{=}SG} \limp S \\
A_{aime_2} &: NP_{NUM{=}SG} \limp NP_{NUM{=}PL} \limp S \\
A_{qui_1} &: (NP_{NUM{=}SG} \limp S) \limp N_{NUM{=}SG} \limp N_{NUM{=}SG} \\
A_{qui_2} &: (NP_{NUM{=}PL} \limp S) \limp N_{NUM{=}PL} \limp N_{NUM{=}PL}
\end{align*}

The types for nouns and determiners are quite predictable. For the transitive
verb \emph{aime}, we see that we need a second type to account for the fact
that \emph{aime} does not care about the number of the object. Finally, the
relativizer \emph{qui} enforces that the number of the trace $NP$ in the
relative clause must be the same as the number of the $N$ being modified.

The lexicon $\mathcal{L}_{Agr}$ follows exactly the same schema as that of
$\mathcal{L}_{Neg}$.

\subsection{Discussion}
